An exploratory analysis of changes in trip-making frequency by mode during the COVID-19 emergency in Bangladesh

### 2020-10-06 Frequency-of-Travel-by-Mode-COVID-19-Bangladesh/Frequency-of-Travel-by-Mode-COVID-19-Bangladesh.tex
v1 Paper submitted to Transport Findings

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

@@

  \begin{abstract}
  The COVID-19 pandemic has had a profound impact on mobility in every
  country and region around the world. Recent research provides
  illuminates the nature and magnitude of the changes in mobility, but the
  evidence is still scant in developing countries. The objective of this
  paper is to present an exploratory analysis of the changes in the
  frequency of trip-making by mode during the COVID-19 emergency in
  Bangladesh. The results confirm an overall loss of mobility in the form
  of reduced trip-making frequency by all modes, but the changes are not
  uniform across modes.
  \end{abstract}

@@

The spread of the COVID-19 pandemic has led to limitations to movement
in many countries and regions, either because of lock-down policies or
self-censoring by segments of the public. The magnitude of changes in
mobility has been studied by recent research, including DeWeese et al.
(2020) and Molloy et al. (2020). While the evidence available indicates
that overall there was a reduction in mobility in much of the world, the
changes were uneven depending on the mode of transportation or the
purpose of the trip (see Lock, 2020; Paez, 2020). Alas, with few
exceptions evidence remains more spotty for developing countries, most
of which have large populations segments that are less able to absorb
losses in mobility (e.g., Astroza et al., 2020; Huynh, 2020; Saha et
al., 2020).

@@

This paper is a reproducible research document (see Brunsdon and Comber,
2020); the code and data necessary to replicate the tables and figures
are available in a public repository\footnote{\url{place.holder}}

@@

Data used for this paper come from a survey conducted during
\textbf{DATES} in \textbf{regions/cities} in Bangladesh. \textbf{BRIEFLY
ADD DETAILS ABOUT SURVEY}.

The survey asked respondents to self-report their trip-making frequency
by eight modes of transportation, namely \textbf{car},
\textbf{ridesharing} (e.g., Uber, Pathao), \textbf{rickshaw},
\textbf{cng auto-rickshaw} (a rickshaw-like vehicle powered by
compressed natural gas), \textbf{bus}, motorcycle/scooter (hereafter
just \textbf{motorcycle}), \textbf{walking}, and \textbf{bicycle} (there
was an additional catch-all category \textbf{other} which we ignore
here). Participants in the survey used the following levels to report
their frequency of traveling by each mode both before and during
COVID-19: \emph{Never}, \emph{Rarely}, \emph{Once a week}, \emph{2-3
trips per week}, \emph{4-5 trips per week}, \emph{Almost daily}. There
are \(n=800\) responses in the data set.

To describe changes in the frequency of travel by mode in the transition
to the pandemic, we use well-established exploratory data analysis (EDA)
techniques.

@@

Figure \ref{fig:column-plot-cases} shows the number of responses (out of
800) in each trip-making frequency class by mode of transportation. The
white bars and gray bars are for travel before and during the pandemic,
respectively. Considering travel before the pandemic, travel by
rickshaw, bus were relatively common for many respondents (few
respondents reported \emph{never} using these modes). The mode most
commonly used on a quotidian basis was walk. In contrast, respondents
reported less frequent travel by car, rideshare services, cng
auto-rickshaw, motorcycle, and bicycle. During the pandemic we see that
while there were reductions in mobility by car, motorcycle, and bicycle
(with more respondents reporting never traveling by these modes), the
changes were relatively minor. The frequency of trip-making by other
modes changed more noticeably: the frequency of travel by rideshare
services, rickshaw, cng auto-rickshaw, and bus collapsed, with vastly
more respondents reporting never using these modes during the pandemic
than before. The frequency of walking trips also decreased (fewer
respondents report walking almost daily), but the reductions in mobility
were not so heavily concentrated at the bottom of the scale.

@@

@@

\begin{table}

\caption{\label{tab:table-transitions}\label{tab:cross-tabulation} Cross-tabulation of counts of travel frequency by mode, before and during COVID-19}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrrr}
\toprule
  & Never & Rarely & Once a week & 2-3 per week & 4-5 per week & Almost daily\\
\midrule
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{car}}\\
\hspace{1em}Never & 267 & 20 & 5 & 1 & 1 & 2\\
\hspace{1em}Rarely & 142 & 99 & 21 & 12 & 9 & 3\\
\hspace{1em}Once a week & 11 & 23 & 19 & 7 & 2 & 3\\
\hspace{1em}2-3 per week & 10 & 23 & 7 & 6 & 8 & 2\\
\hspace{1em}4-5 per week & 4 & 9 & 5 & 6 & 12 & 4\\
\hspace{1em}Almost daily & 6 & 15 & 6 & 10 & 0 & 20\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{rideshare}}\\
\hspace{1em}Never & 254 & 13 & 6 & 4 & 0 & 0\\
\hspace{1em}Rarely & 192 & 44 & 12 & 1 & 1 & 1\\
\hspace{1em}Once a week & 60 & 20 & 31 & 3 & 1 & 0\\
\hspace{1em}2-3 per week & 49 & 12 & 9 & 7 & 2 & 0\\
\hspace{1em}4-5 per week & 28 & 10 & 3 & 5 & 6 & 2\\
\hspace{1em}Almost daily & 14 & 3 & 4 & 1 & 0 & 2\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{rickshaw}}\\
\hspace{1em}Never & 51 & 9 & 6 & 0 & 1 & 0\\
\hspace{1em}Rarely & 64 & 70 & 18 & 4 & 3 & 1\\
\hspace{1em}Once a week & 36 & 38 & 45 & 9 & 3 & 3\\
\hspace{1em}2-3 per week & 34 & 36 & 18 & 29 & 6 & 4\\
\hspace{1em}4-5 per week & 21 & 30 & 7 & 16 & 15 & 8\\
\hspace{1em}Almost daily & 45 & 65 & 30 & 18 & 16 & 41\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{cng auto-rickshaw}}\\
\hspace{1em}Never & 176 & 14 & 3 & 1 & 0 & 0\\
\hspace{1em}Rarely & 213 & 95 & 14 & 4 & 1 & 3\\
\hspace{1em}Once a week & 36 & 31 & 38 & 9 & 2 & 3\\
\hspace{1em}2-3 per week & 29 & 18 & 16 & 14 & 8 & 3\\
\hspace{1em}4-5 per week & 13 & 6 & 5 & 9 & 14 & 1\\
\hspace{1em}Almost daily & 8 & 7 & 2 & 0 & 0 & 4\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{bus}}\\
\hspace{1em}Never & 116 & 9 & 3 & 0 & 0 & 0\\
\hspace{1em}Rarely & 155 & 46 & 9 & 0 & 0 & 1\\
\hspace{1em}Once a week & 73 & 32 & 22 & 4 & 0 & 0\\
\hspace{1em}2-3 per week & 43 & 19 & 11 & 7 & 5 & 0\\
\hspace{1em}4-5 per week & 60 & 6 & 5 & 4 & 12 & 1\\
\hspace{1em}Almost daily & 103 & 21 & 10 & 11 & 2 & 10\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{motorcycle}}\\
\hspace{1em}Never & 467 & 22 & 12 & 3 & 1 & 8\\
\hspace{1em}Rarely & 44 & 38 & 13 & 10 & 0 & 4\\
\hspace{1em}Once a week & 11 & 11 & 34 & 3 & 3 & 3\\
\hspace{1em}2-3 per week & 13 & 6 & 6 & 15 & 4 & 2\\
\hspace{1em}4-5 per week & 6 & 4 & 0 & 4 & 14 & 3\\
\hspace{1em}Almost daily & 1 & 3 & 2 & 4 & 3 & 23\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{walk}}\\
\hspace{1em}Never & 17 & 7 & 3 & 4 & 0 & 2\\
\hspace{1em}Rarely & 26 & 53 & 13 & 6 & 2 & 4\\
\hspace{1em}Once a week & 12 & 19 & 54 & 9 & 4 & 2\\
\hspace{1em}2-3 per week & 10 & 14 & 12 & 26 & 13 & 4\\
\hspace{1em}4-5 per week & 7 & 3 & 7 & 11 & 16 & 8\\
\hspace{1em}Almost daily & 46 & 73 & 40 & 49 & 27 & 197\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{bicycle}}\\
\hspace{1em}Never & 504 & 22 & 7 & 4 & 2 & 4\\
\hspace{1em}Rarely & 59 & 43 & 20 & 4 & 2 & 4\\
\hspace{1em}Once a week & 14 & 13 & 21 & 3 & 2 & 1\\
\hspace{1em}2-3 per week & 3 & 4 & 5 & 7 & 2 & 0\\
\hspace{1em}4-5 per week & 2 & 2 & 4 & 1 & 9 & 3\\
\hspace{1em}Almost daily & 7 & 1 & 1 & 1 & 3 & 16\\
\bottomrule
\end{tabular}}
\end{table}

To more easily understand the transitions towards different trip-making
frequencies, including possible adaptations by mode, we convert the
cross-tabulations to probability transition matrices, which we then
visualize using circular plots.

Figures \ref{fig:circular-plot-1} to \ref{fig:circular-plot-4} present
these plots. Each of the trip-making frequency sectors on the left
hemisphere of the circle represent 100\% of responses \emph{before} the
pandemic. The size of the links is proportional to the probability
\(p_{ij}\) of transitioning from frequency class \(i\) before to the
right hemisphere are proportional to the transition probabilities to
each frequency \emph{during} the pandemic. There are three transparency
levels for the links: solid colors are for \(p_{ij}>2/3\), intermediate
transparency is for \(1/3 < p_{ij} \le 2/3\), and the more transparent
links are for \(p_{ij}\le 1/3\).

From Figure \ref{fig:circular-plot-1} we see that the probability of
traveling less frequently for those who frequently traveled by car
before is high, but their probability of not using this mode at all
during the pandemic is quite small. The probability of traveling more
frequently by car for those who originally never or rarely used this
mode remained is low. In contrast, we see that the probabilities of
never ridesharing during the pandemic are high irrespective of the
initial level of use of this mode of transportation.

The probabilities of change in trip frequency by rickshaw and and cng
auto-rickshaw are similar (see Figure \ref{fig:circular-plot-2}),
although the probabilities of being less mobile by cng auto-rickshaw are
greater: between 1/3 and 2/3 of respondents who used this mode almost
daily, stopped using it during the pandemic. Very rarely there was an
increase in mobility by these modes.

After ridesharers, bus users (Figure \ref{fig:circular-plot-3}) were the
most likely to stop using it, irrespective of their initial level of use
of this mode, while the probabilities of using it \emph{more} frequently
are extremely small.

@@

@@

>>>>

% Pandoc header
\usepackage{float} \floatplacement{figure}{H}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}

@@

  \begin{abstract}
  The COVID-19 pandemic has had a profound impact on mobility in every
  country and region around the world. Recent helps to illuminate some of
  the dimensions of change - however, the evidence is still scant in
  developing countries. The objective of this paper is to present an
  exploratory analysis of the changes in the frequency of trip-making by
  mode during the COVID-19 emergency in Bangladesh. The results confirm an
  overall loss of mobility in the form of reduced trip-making frequency by
  all modes, but the changes are not uniform across modes. In particular,
  there was a greater loss of mobility for bus, rickshaw, and CNG
  auto-rickshaw.
  \end{abstract}

@@

The spread of the COVID-19 pandemic has led to limitations to movement
in many countries and regions, either because of lock-down policies or
self-censoring by segments of the public. The magnitude of changes in
mobility has been studied by recent research, including DeWeese et al.
(2020) and Molloy et al. (2020). While the evidence available indicates
that although the overall mobility has reduced in much of the world, the
changes were uneven depending on the mode of transportation or the
purpose of the trip (see Lock, 2020; Paez, 2020). Unfortunately, with
few exceptions, evidence remains more spotty for developing countries,
most of which have large population segments that are less able to
absorb losses in mobility (e.g., Astroza et al., 2020; Huynh, 2020; Saha
et al., 2020).

@@

This paper is a reproducible research document (see Brunsdon and Comber,
2020); the code and data necessary to reproduce the tables and figures
are available in a public
repository\footnote{\url{https://github.com/paezha/Frequency-of-Travel-by-Mode-COVID-19-Bangladesh}}

@@

Data used for this paper come from an online survey, ``Exploring the
potential of travel mode change behavior in the post-lock-down and
post-pandemic (COVID-19) period'' conducted during July - August 2020 in
Bangladesh. The survey was available both in Bengali and English and
disseminated through various electronic means along with a widespread
social media campaign and promotion in various social media sites and
groups. In addition to socio-economic and demographic information of the
respondents, the survey covered different aspects of travel such as
travel behavior before and during COVID-19, knowledge related to
COVID-19, and opinions and perceptions regarding travel behavior after
lock-down and post-pandemic situation.

For the purpose of this paper, we use two questions that provide
information about trip-making frequency by eight modes of
transportation. The modes are \textbf{car}, \textbf{ridesharing} (e.g.,
Uber, Pathao), \textbf{rickshaw}, \textbf{CNG auto-rickshaw} (a
rickshaw-like vehicle powered by compressed natural gas), \textbf{bus},
motorcycle/scooter (hereafter just \textbf{motorcycle}),
\textbf{walking}, and \textbf{bicycle} (there was an additional
catch-all category \textbf{other} which we ignore here). Participants in
the survey used the following levels to report their frequency of
traveling by each mode both before and during COVID-19: \emph{Never},
\emph{Rarely}, \emph{Once a week}, \emph{2-3 trips per week}, \emph{4-5
trips per week}, \emph{Almost daily}. There are \(n=800\) responses in
the data set.

To describe changes in travel frequency by mode in the transition to the
pandemic, we use well-established exploratory data analysis (EDA)
techniques.

@@

Figure \ref{fig:column-plot-cases} shows the number of responses (out of
800) in each trip-making frequency class by mode of transportation. The
white bars and gray bars are for travel before and during the pandemic,
respectively. Considering travel before the pandemic, travel by
rickshaw, and bus were relatively common for many respondents (few
respondents reported \emph{never} using these modes). The mode most
commonly used on a quotidian basis was walk. In contrast, respondents
reported less frequent travel by car, ridesharing services, CNG
auto-rickshaw, motorcycle, and bicycle. During the pandemic, we see that
while there were reductions in mobility by car, motorcycle, and bicycle
(with more respondents reporting never traveling by these modes), the
changes were relatively minor.

@@

\begin{table}[H]

\caption{\label{tab:table-transitions}\label{tab:cross-tabulation} Cross-tabulation of counts of travel frequency by mode, before and during COVID-19}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrrr}
\toprule
  & Never & Rarely & Once a week & 2-3 per week & 4-5 per week & Almost daily\\
\midrule
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{car}}\\
\hspace{1em}Never & 267 & 20 & 5 & 1 & 1 & 2\\
\hspace{1em}Rarely & 142 & 99 & 21 & 12 & 9 & 3\\
\hspace{1em}Once a week & 11 & 23 & 19 & 7 & 2 & 3\\
\hspace{1em}2-3 per week & 10 & 23 & 7 & 6 & 8 & 2\\
\hspace{1em}4-5 per week & 4 & 9 & 5 & 6 & 12 & 4\\
\hspace{1em}Almost daily & 6 & 15 & 6 & 10 & 0 & 20\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{rideshare}}\\
\hspace{1em}Never & 254 & 13 & 6 & 4 & 0 & 0\\
\hspace{1em}Rarely & 192 & 44 & 12 & 1 & 1 & 1\\
\hspace{1em}Once a week & 60 & 20 & 31 & 3 & 1 & 0\\
\hspace{1em}2-3 per week & 49 & 12 & 9 & 7 & 2 & 0\\
\hspace{1em}4-5 per week & 28 & 10 & 3 & 5 & 6 & 2\\
\hspace{1em}Almost daily & 14 & 3 & 4 & 1 & 0 & 2\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{rickshaw}}\\
\hspace{1em}Never & 51 & 9 & 6 & 0 & 1 & 0\\
\hspace{1em}Rarely & 64 & 70 & 18 & 4 & 3 & 1\\
\hspace{1em}Once a week & 36 & 38 & 45 & 9 & 3 & 3\\
\hspace{1em}2-3 per week & 34 & 36 & 18 & 29 & 6 & 4\\
\hspace{1em}4-5 per week & 21 & 30 & 7 & 16 & 15 & 8\\
\hspace{1em}Almost daily & 45 & 65 & 30 & 18 & 16 & 41\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{CNG auto-rickshaw}}\\
\hspace{1em}Never & 176 & 14 & 3 & 1 & 0 & 0\\
\hspace{1em}Rarely & 213 & 95 & 14 & 4 & 1 & 3\\
\hspace{1em}Once a week & 36 & 31 & 38 & 9 & 2 & 3\\
\hspace{1em}2-3 per week & 29 & 18 & 16 & 14 & 8 & 3\\
\hspace{1em}4-5 per week & 13 & 6 & 5 & 9 & 14 & 1\\
\hspace{1em}Almost daily & 8 & 7 & 2 & 0 & 0 & 4\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{bus}}\\
\hspace{1em}Never & 116 & 9 & 3 & 0 & 0 & 0\\
\hspace{1em}Rarely & 155 & 46 & 9 & 0 & 0 & 1\\
\hspace{1em}Once a week & 73 & 32 & 22 & 4 & 0 & 0\\
\hspace{1em}2-3 per week & 43 & 19 & 11 & 7 & 5 & 0\\
\hspace{1em}4-5 per week & 60 & 6 & 5 & 4 & 12 & 1\\
\hspace{1em}Almost daily & 103 & 21 & 10 & 11 & 2 & 10\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{motorcycle}}\\
\hspace{1em}Never & 467 & 22 & 12 & 3 & 1 & 8\\
\hspace{1em}Rarely & 44 & 38 & 13 & 10 & 0 & 4\\
\hspace{1em}Once a week & 11 & 11 & 34 & 3 & 3 & 3\\
\hspace{1em}2-3 per week & 13 & 6 & 6 & 15 & 4 & 2\\
\hspace{1em}4-5 per week & 6 & 4 & 0 & 4 & 14 & 3\\
\hspace{1em}Almost daily & 1 & 3 & 2 & 4 & 3 & 23\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{walk}}\\
\hspace{1em}Never & 17 & 7 & 3 & 4 & 0 & 2\\
\hspace{1em}Rarely & 26 & 53 & 13 & 6 & 2 & 4\\
\hspace{1em}Once a week & 12 & 19 & 54 & 9 & 4 & 2\\
\hspace{1em}2-3 per week & 10 & 14 & 12 & 26 & 13 & 4\\
\hspace{1em}4-5 per week & 7 & 3 & 7 & 11 & 16 & 8\\
\hspace{1em}Almost daily & 46 & 73 & 40 & 49 & 27 & 197\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{bicycle}}\\
\hspace{1em}Never & 504 & 22 & 7 & 4 & 2 & 4\\
\hspace{1em}Rarely & 59 & 43 & 20 & 4 & 2 & 4\\
\hspace{1em}Once a week & 14 & 13 & 21 & 3 & 2 & 1\\
\hspace{1em}2-3 per week & 3 & 4 & 5 & 7 & 2 & 0\\
\hspace{1em}4-5 per week & 2 & 2 & 4 & 1 & 9 & 3\\
\hspace{1em}Almost daily & 7 & 1 & 1 & 1 & 3 & 16\\
\bottomrule
\end{tabular}}
\end{table}

The frequency of trip-making by other modes changed more noticeably: the
frequency of travel by ridesharing services, rickshaw, CNG
auto-rickshaw, and bus collapsed, with vastly more respondents reporting
never using these modes during the pandemic than before. The frequency
of walking trips also decreased (fewer respondents report walking almost
daily), but the reductions in mobility were not so heavily concentrated
at the bottom of the scale.

@@

To more clearly understand the transitions towards different trip-making
frequencies, including possible adaptations, we convert the
cross-tabulations to probability transition matrices, which we then
visualize using circular plots.

Figures \ref{fig:circular-plot-1} to \ref{fig:circular-plot-4} present
these plots. Each of the trip-making frequency sectors on the left
hemisphere of the circle represents 100\% of responses \emph{before} the
pandemic. The links' size is proportional to the probability \(p_{ij}\)
of transitioning from frequency class \(i\) before to the right
hemisphere are proportional to the transition probabilities to each
frequency \emph{during} the pandemic. There are three transparency
levels for the links: solid colors are for \(p_{ij}>2/3\), intermediate
transparency is for \(1/3 < p_{ij} \le 2/3\), and the more transparent
links are for \(p_{ij}\le 1/3\).

From Figure \ref{fig:circular-plot-1}, we see that the probability of
traveling less by car for those who initially used this mode frequently
is high, but their probability of not using this mode at all during the
pandemic is quite small. In other words, there was a decline in use, but
not complete discontinuation of this mode. The probability of traveling
more frequently by car for those who originally never or rarely used
this mode remained is low. In contrast, we see that the probabilities of
never ridesharing during the pandemic are high irrespective of the
initial level of use of this mode of transportation, and the probability
of using this mode more are fairly small.

@@

After rideshare, the use of bus had the largest probabilities of
discontinuation (Figure \ref{fig:circular-plot-3}). We see that even
respondents who used this mode almost daily before the pandemic had
close to 66\% chances of never using it during the contingency.
Similarly large losses of mobility by bus were observed for travelers
who used this mode less frequently before the pandemic. These losses
were not offset to any appreciable degree by the probability of some
users turning to this mode during the pandemic.

@@

Trip-making by motorcycle (Figure \ref{fig:circular-plot-3}) also
declined for all but the most frequent users pre-COVID-19, and the
probability of adopting this mode during COVID-19 was almost null.
Contrast this to the case of walking (Figure \ref{fig:circular-plot-4}),
where a sizable number of respondents started walking during the
pandemic, even if only rarely. Close to 50\% of respondents who report
walking almost daily during the pandemic walked less frequently before
the contingency. Bicycle (Figure \ref{fig:circular-plot-4}), like
motorcycle, saw relatively fewer respondents adopting it during the
pandemic, and many travelers who did use the mode rarely, once a week,
or even 2-3 times per week stopped cycling during the pandemic.
